NumPy---Array Programming in Python

### 2020-01-30 body.tex
Thoughts about expanding the prolif section based on Jarrods ideas

NumPy provides in-memory, multidimensional, homogeneously typed
(i.e., single pointer and strided) arrays on CPUs.
It supports many types of CPUs (e.g., POWER and ARM) with similar
performance as native C code.
% a brief word about accelerators perhaps?
% end with how this fits the bill for a huge number of users

While NumPy suits most user's needs, it cannot support every specialized need
of the community; due to its in-memory data model, it is limited in its ability
to work with very large datasets, datasets split across multiple systems, and
computations that require graphics processing units (GPUs), tensor processing
units (TPUs) or other specialized hardware for performance reasons.

% maybe mention that there are typically utilities
% to convert to and from numpy arrays
% and why (presumably to get access to ecosystem)
The community's efforts to fill the resultant gaps led to a
proliferation of array implementations. Each deep learning framework created
its own arrays; PyTorch \cite{NEURIPS2019_9015},
Tensorflow \cite{abadi2016tensorflow}, Apache MXNet \cite{chen2015mxnet},
and JAX \cite{jax2018github} arrays all have the
capability to run on CPUs and GPUs, in a distributed fashion, and in a delayed
execution mode to allow for performance optimizations.  SciPy and Sparse both
provide sparse arrays---which typically contain few non-zero values and store
only those in memory for efficiency.
In addition there are projects that build on top of NumPy arrays as a data
container and \textit{extend} its capabilities.  Distributed arrays are
made possible that way by Dask, and labeled arrays---referring to dimensions of
an array by name rather than by index for clarity, compare \code{x[:,~1]} vs.
\code{x.loc[:,~'time']}---by Xarray \cite{hoyer2017xarray}.

@@

Over time the role of the NumPy has evolved.
It began as an attempt to add an array object to Python
and became the foundation of a rich ecosystem of tools.
Now a large amount of scientific work depends on NumPy
being correct, fast, and stable.
Yet, it is no longer just the foundational array library underlying the
scientific Python ecosystem; it is also the standard API for tensor
computation and a central coordinating mechanism between array types and
technologies in Python.
NumPy continues to evolve.
For instance, the NumPy data type system is currently being redesigned to make it
easier to implement data types both in Python and C, and to better handle
richer data types including physical units \cite{astropy,Goldbaum2018,pint},
geometrical objects \cite{pygeos}, and missing values.

>>>>

NumPy provides in-memory, multidimensional, homogeneously typed
(i.e., single pointer and strided) arrays on CPUs.
It runs on machines ranging from embedded devices up to the worlds
largest supercomputers [citation needed] and provides a performance
which often approaches compiled languages.
When algorithms do not fit the array programming approach,
or require a faster implementation,
a number of tools (e.g. Cython, Numba, Pythran, ...)
are available to transparently accelerate crucial parts.
These tools allow for a high level array orientated code base without sacrificing
the speed expected from a compiled language.
In many cases fast algorithms are implemented by libraries in the scientific
python ecosystem so that most user programs are sufficiently fast being
written in Python and NumPy.
% smooth experience/intengration from high level <-> low level?

% a brief word about accelerators perhaps?
% end with how this fits the bill for a huge number of users

NumPy suits most user's needs, but it cannot support every specialized need
of the community; due to its in-memory data model, it is limited in its ability
to work with very large datasets, datasets split across multiple systems, and
computations that require graphics processing units (GPUs), tensor processing
units (TPUs) or other specialized hardware for performance reasons.
However, both distributed data and the parallel execution of GPUs and TPUs
map well to the paradigm of array orientated programming.
This leads to a gap between available modern hardware architecture and the
necessary tools to leverage their computational power.

% maybe mention that there are typically utilities
% to convert to and from numpy arrays
% and why (presumably to get access to ecosystem)
The community's efforts to fill the resultant gaps led to a
proliferation of new array implementations. Each deep learning framework created
its own arrays; PyTorch \cite{NEURIPS2019_9015},
Tensorflow \cite{abadi2016tensorflow}, Apache MXNet \cite{chen2015mxnet},
and JAX \cite{jax2018github} arrays all have the
capability to run on CPUs and GPUs, in a distributed fashion, and in a delayed
execution mode to allow for performance optimizations.  SciPy and Sparse both
provide sparse arrays---which typically contain few non-zero values and store
only those in memory for efficiency.
In addition there are projects that build on top of NumPy arrays as a data
container and \textit{extend} its capabilities.  Distributed arrays are
made possible that way by Dask, and labeled arrays---referring to dimensions of
an array by name rather than by index for clarity, compare \code{x[:,~1]} vs.
\code{x.loc[:,~'time']}---by Xarray \cite{hoyer2017xarray}.

@@

Over time the role of the NumPy has evolved.
It began as an attempt to add an array object to Python
and became the foundation of a rich ecosystem of tools.
Now a large amount of scientific work depends on NumPy
being correct, fast, and stable.
Yet, it is no longer just the foundational array library underlying the
scientific Python ecosystem; it is also the standard API for tensor
computation and a central coordinating mechanism between array types and
technologies in Python.
NumPy continues to evolve.
For instance, the NumPy data type system is currently being redesigned to make it
easier to implement data types both in Python and C, and to better handle
richer data types including physical units \cite{astropy,Goldbaum2018,pint},
geometrical objects \cite{pygeos}, and missing values.
% Maybe add a sentence in the direction of:
% We hope/expect, that this will allow the proliferation of rich datatypes
% which in turn can propel the whole ecosystem forward? (too much :)?
% As a part of other changes/developements...)
% Discussed with Jarrod, if it fits to get back to protocols, it is fine
% to call DTypes a (rich) protocol.
